---
title: "Reproducible Research: Peer Assessment 2"
output: 
  html_document:
    keep_md: true
---

## Synopsis

Analysis of the greatest weather threats to public health and economy of the
United States.

## Environment information

### Used libraries

```{r echo=TRUE,message=FALSE}
library(dplyr)
library(R.oo)
library(ggplot2)
library(reshape2)
```

### Session information

This analysis war performed using the software listed below.

```{r echo=TRUE}
sessionInfo()
```


## Data Processing

Data may be obtained from the [National Weather Service](http://www.weather.gov/)
[Storm Data Documentation](http://www.ncdc.noaa.gov/stormevents/) although data 
data used for the current analysis was provided from:

* [StormData.csv.bz2](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)

There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined.

* National Weather Service [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)
* National Climatic Data Center Storm Events [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

### Loading and preprocessing the data

Raw data is extrated from it's [original distribution file](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2).

```{r echo=TRUE, cache=TRUE}
raw_data <- read.csv(bzfile("StormData.csv.bz2"), na.strings="", stringsAsFactors=FALSE)
```

The data consists of `r dim(raw_data)[1]` rows of `r dim(raw_data)[2]`
variables.

First show some exploratory information about the variables used in the analysis.

```{r echo=TRUE}
# Summary of event types and count per year
yeve <- raw_data %>%
    mutate(year=as.integer(format(as.Date(raw_data$BGN_DATE,"%m/%d/%Y"),"%Y"))) %>%
    group_by(year,EVTYPE) %>%
    summarise(cnt=n())  %>% summarise(count=sum(cnt),n=n())
par(mfrow = c(1,2))
plot(n ~ year, yeve, main="Different event types reported by year")
plot(count ~ year, yeve, main="Reported events per year")
```

Events in the database start in the year `r min(yeve$year)` to `r max(yeve$year)`.

Before 1993 only few types of events were reported, also the quantity of events 
reported is much higher after 1993.

The event type (EVTYPE) has non consistent values, a simple aproach to clean
this data is to clean extra whitespace and make it all uppercase.

```{r echo=TRUE}
raw_evType_summary = raw_data %>% select(EVTYPE) %>% group_by(EVTYPE) %>% summarise()
clean_evType = data.frame(EVTYPE=as.factor(toupper(trim(raw_data$EVTYPE))))
clean_evType_summary = clean_evType %>% select(EVTYPE) %>% group_by(EVTYPE) %>% summarise()
```

This simple cleanup of the event type colapsed the numbert of different events
from `r dim(raw_evType_summary)[1]` to `r dim(clean_evType_summary)[1]`.

Property and crop damage are expresed as numbers with magnitude factors, this
factors should be K, M, B standing for thousands, millions and billions
respectively, but the file has errors shown in the folowing summary:

```{r echo=TRUE}
summary(as.factor(raw_data$PROPDMGEXP))
```

It will be assumed that missing factors account for K, that upper and lower case
factors of the same letter have the same value and, that all extrange factor
names (which have very little samples) are thousands. The function used to do
this conversion is shown below.

```{r echo=TRUE}
parse_exp <- function(e) {
  if (is.na(e)) 1e3
  else if (tolower(e) == "k") 1e3
  else if (tolower(e) == "m") 1e6
  else if (tolower(e) == "b") 1e9
  else 1e3
}
```

Clean up and filter the data, just keep the rows and variables needed for
the analysis.

```{r echo=TRUE, cache=TRUE}
data <- data.frame(
  date = as.Date(raw_data$BGN_DATE,"%m/%d/%Y"),
  evType = as.factor(toupper(trim(raw_data$EVTYPE))),
  fatalities = as.integer(raw_data$FATALITIES),
  injuries = as.integer(raw_data$INJURIES),
  property = raw_data$PROPDMG * sapply(raw_data$PROPDMGEXP, parse_exp),
  crop = raw_data$CROPDMG * sapply(raw_data$CROPDMGEXP, parse_exp)
  )
```

To avoid bias, only years after 1993 will be considered for the analysis.

```{r echo=TRUE, cache=TRUE}
data <- filter(data, format(date,"%Y") > "1993")
```

Compute top causes of health and property damage.

```{r echo=TRUE}
sum_by_evType <- data %>% group_by(evType) %>% 
    summarise(fatalities=sum(fatalities),injuries=sum(injuries),property=sum(property),crop=sum(crop))

top_num <- 20
top_health <- sum_by_evType %>%  top_n(top_num,fatalities)
top_property <- sum_by_evType %>% top_n(top_num,property+crop)
```


## Results

Top health threads

```{r echo=TRUE}
f <- top_health %>% select(evType,fatalities,injuries)
f$evType <- reorder(f$evType, f$injuries+f$fatalities)
f <- melt(f,id.vars = "evType")
ggplot(data=f, aes(evType, value)) + geom_bar(stat="identity",aes(fill=variable)) +
    coord_flip() + xlab('')+ ylab('Fatalities count') +
    ggtitle("Top 20 Causes of health thread")
```

Top property threads

```{r echo=TRUE}
f <- top_property %>% select(evType,property,crop)
f$evType <- reorder(f$evType, f$property+f$crop)
f <- melt(f,id.vars = "evType")
f$value <- f$value / 1e9
ggplot(data=f, aes(evType, value)) + geom_bar(stat="identity",aes(fill=variable)) +
    coord_flip() + xlab('')+ ylab('Looses in billions of USD') +
    ggtitle("Top 20 Causes of property damage")
```
